Delayed dichromatism as a convenient tool to disentangle the effects of survival and productivity on the population dynamics in waterfowl

### 2024-05-27 Code/Draft_ruddy_duck_vital_rates.Rmd
Before instatext

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Delayed dichromatism as a convenient tool to disentangle the effects of survival and productivity on population growth in waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]
- Jean-François Maillard^[Office Français de la Biodiversité, France - jean-francois.maillard@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]
date: "`r Sys.setlocale('LC_TIME', 'English'); format(Sys.time(), '%d %B, %Y')`"
abstract: |
  Monitoring the number of individuals is by far the most popular strategy for studying the environmental factors that rule population dynamics and for measuring the efficacy of management actions aimed at population recovery, control or eradication. Monitoring the number of individuals is unfortunately inefficient in assessing the mechanisms underlying demographic processes and, more specifically, the extent to which population growth rate is affected by changes in adult survival rather than variations in reproductive parameters. Common methods to inform survival and recruitment, such as capture-mark-recapture and game-hunting bag, rely on catching animals, which has three main disadvantages. First, the selectivity of capture methods can lead to a biased representation of the underlying population structure. Second, catching and releasing animals is quite invasive, and causes stress, thus posing an additional threat to often endangered species. Third, release may simply be impossible, especially for invasive alien species that represent a threat to native taxa. In many waterfowl species, sexual dichromatism is observed in adults, while immatures of both sexes display a cover-up plumage similar to that of adult females. Using the example of non-native populations of Ruddy duck in Western Europe, a species that exhibits this pattern, we present a non-invasive method based on count data to estimate the respective contribution of recruitment and adult survival to the population growth rate. Using counts, we illustrate how changes in the apparent sex ratio (due to female-like young males gradually acquiring the plumage of adult male) can be exploited to partition population growth rate into adult survival and recruitment, and how this contributes significantly to understanding the demographic impact of two different eradication strategies.

  **Keywords**: population dynamics - survival - reproductive success - fecundity - vital rate - age ratio - intrinsic growth rate - invasive species - Oxyura jamaicensis
output: 
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
    bookdown::word_document2:
      toc: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

Assessing population growth rate is a key step towards a better understanding of factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also crucial for measuring the efficacy of management actions potentially undertaken to help population recovery, control, or eradication [@Shea1998]. Among the approaches available to managers for reaching these goals, those relying on the monitoring of the number of individuals, i.e. counts, are by far the most popular [@Rintala2022]. In many cases, these methods even allow the investigation of environmental factors underlying changes in population size and thus help in the implementation of mitigation measures [@Faillettaz2019]. Unfortunately, relying on the monitoring of numbers of individuals alone generally hinders basic understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, based on counts only, it is nearly impossible to assess the extent to which population growth rate is affected by change in survival rather than to variations in reproductive parameters [@Austin2000]. To assess the relative sensitivity of population growth rate to factors affecting survival or reproductive parameters, demographers most often rely on the monitoring of individuals by capture-mark-recapture [@Lebreton1992]. In practice, however, one cannot always rely on the capture and release of individuals owing to the legal status of the species that precludes any additional disturbance, e.g. critically endangered species, or forbid the release of alive individuals, e.g. invasive species or pests. Furthermore, although highly efficient for assessing demographic parameters, CMR methods suffer from potential drawbacks such as being invasive (and may affect behaviour and hence survival or breeding success) and hardly affordable when time and money are limiting. Genetic monitoring is a non-invasive alternative to capture-mark-recapture approaches based on physical captures, but it suffers from being costly, and requires large samples [@Caniglia2011].

Most of the time therefore, managers do the best of a bad job using counts as a viable option for tracking population growth rates and assessing the relevance of management actions. Counting the number of immatures in addition to live adults may enable the estimation of adult survival and recruitment rates, and thus allow the assessment of their relative influence on population growth rate. The adult survival rate is the proportion of adults staying alive over one year. The recruitment rate, also called productivity [e.g. @Hagen2008; @Johnson1987], is the average number of immatures per breeder that reach the sexual maturity, and is thus the product of two main sub-parameters: the fecundity followed by the juvenile survival [@Arnold2018; @Etterson2011]. Alternatively, assessing the sex and age structure in hunting bags has been used to infer the role of decreasing reproductive success in population declines in a number of games species including ducks and geese but suffers from intractable biases [@Fox2018]. Of course, this approach is not suited for protected / endangered species.

Here we exemplify how, in dimorphic species, delayed maturity of males can be used for estimating adult survival and recruitment rate by differentiating male-like from female-like individuals in repeated winter counts. We used two non-native Ruddy duck (*Oxyura jamaicensis*) populations introduced in Europe as a study model. As numerous duck species, Ruddy duck is dimorphic, with newly born males looking like females until the pre-nuptial moult, which occurs in late winter. As a result of this late prenuptial moult of immatures, the apparent proportion of males increases over the course of the wintering season. These changes in apparent proportions of males during this period are therefore directly related to the proportion of immatures into the populations and thus to the reproductive success of the previous breeding season, and the recruitment rate of the next one. We estimated adult survival and recruitment rates and assessed the impact of two different eradication strategies deployed in Great Britain and in France, respectively.

@@

The Ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in the 1940’s [@GutierrezExposito2020], a feral population began to establish in Great Britain with the first observed reproductive attempts in the wild in the 60s (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5,000 individuals in the early 2000s. By the end of the 80s, a feral population began to establish also on the continent, especially in France, supposedly owing to the arrival of individuals born in Great Britain. But contrary to what was observed in Great Britain, the Ruddy ducks did not spread much in France, with the vast majority of observations and breeding attempts clustering in the North-West of the territory. In France, during winter, almost no Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which greatly facilitates the monitoring of this population. Ruddy ducks are considered as a major threat to the endangered native White-headed duck (*Oxyura leucocephala*) of the South-Western Mediterranean, because of hybridization and thus an elevated risk for genetic pollution and genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, eradication measures were implemented in the late 1990's both in Great Britain and in France [@GutierrezExposito2020], (Figure \@ref(fig:count)), and were followed by a European Ruddy duck eradication plan in 1999 [@Hughes1999].

The trends and status of populations of both White-headed duck and Ruddy duck are being assessed through population censuses only. It is thus impossible to assess the factors that impair recovery of the former and the efficacy of the eradication program for the latter. In particular, based on censuses alone, one cannot assess of the relative effects of changes in adult survival and breeding success or recruitment rates on population growth rates which are prerequisites to identifying limiting factors.

The release of non-native species in the wild, including Ruddy ducks, is forbidden in countries of the European Union, and the status of White-headed duck populations is highly unfavourable. The use of capture-mark-recapture to monitor populations of these species is thus impossible. As a result, the effectiveness of the eradication of Ruddy ducks and management actions intended to favour the recovery of White-headed ducks were essentially assessed through censuses. Censuses, however, prevented assessment of the relative effects or changes in survival and recruitment rates on population growth rate, which is a key towards a proper understanding of the efficiency of management actions.

```{r count, fig.cap = "(ref:count)"}
knitr::include_graphics("../Output/plot_paper_2_mod.png")
```

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism, thereby, young males acquire the typical colourful plumage of their kind (and thus can be distinguished from females) over the course of the interbreeding season, typically by mid-winter for the earliest individuals [@Baldassarre2014]. Delayed dichromatism generally explains the discrepancies in the estimates of the apparent proportions of females in winter vs spring, or between winters counts and hunting bags counts (Figures \@ref(fig:flock) & \@ref(fig:pfem)).

Interestingly, delayed dichromatism allow the estimation of adult survival and recruitment rates by monitoring the seasonal evolution of apparent sex ratio and by making few and reasonable assumptions (see below).

Thus, it is possible to assess the proportion of immatures within a population by monitoring the seasonal evolution of apparent sex ratio and assuming an even sex ratio at birth [@Bellrose1961; @Blums1996], comparable mortality rates between females and males among juveniles [@Ramula2018; @Wood2021], and unchanging adult sex ratio over the monitoring period [@Devineau2010]. Then, both adult survival rate and recruitment rate can be estimated by combining the age ratio thus obtained with reliable counts over two successive winters.

@@

```{r pfem, fig.cap = "(ref:pfem)"}
knitr::include_graphics("../Output/plot_paper_1_saved.png")
```

We monitored the demography of Ruddy duck populations in both Great Britain and France using exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition, counts that distinguished female-like from male-like individuals were performed from 2006 to 2012 in Great Britain, which corresponded to a period of strong decline of the population (Figure \@ref(fig:count)). In France, the monitoring spread over 1999, 2001-2009, and 2012-2019. This population grew rapidly during the first years of the monitoring and then stabilized from 2006 onwards, which corresponds to a period with a high culling effort, which was performed under control of the authorities in both countries. Even if the age ratio in the removals was uncertain before 2009 in France, it is possible to grossly qualify the culling pressure in three categories for both countries: "no culling" before 1999, "low culling" between 1999 and 2005, and "high culling" from 2006 (see Figure \@ref(fig:sup) in Supplementary Materials).

## Vital rates inference from count data

In waterfowl, survival rate is commonly higher in adult males than in adult females [@Baldassarre2014]. Adult sex ratio is then generally biased towards males (Figure \@ref(fig:pfem), @Szekely2014). As a consequence, deducing the proportion of immatures within a population from the observed proportion of adult males is not straightforward because the proportion of adults is not just twice as much as the observed proportion of adult males. There were no counts differentiating male-like from female-like individuals during the breeding period to properly estimate adult sex ratio. But sex identification in adults from removals was available for both Ruddy duck populations. However, the small population size in France prevented from getting precise adult sex ratios. A comparison over months in Great Britain showed no difference in proportion of males in adults, so data from adults collected over the whole year were used to estimate the proportion of males in adults. A comparison of proportion of males in adults among years with more than 500 samples did not exhibit significant interannual differences. Even if the adult sex ratio may vary over a long time range, it is relatively stable over a few years [@Wood2021]. As a consequence, removal data from all adults were pooled to estimate the proportion of males in adults, see Equation \@ref(eq:pm).

Assuming the additive property of the binomial distribution, the proportion of immatures can be deduced from the cumulated counts of male-like individuals in the wintering population, see Equation \@ref(eq:pic). From this proportion and the interannual variation of an abundance index of population size, adult survival and recruitment rate are straightforward, see Equation \@ref(eq:vr). If the absolute value of population size is accessible, numbers of adults and recruits can be estimated, see Equation \@ref(eq:VR).

|Name|Class|Description|
|:---|:----|:----------------------------|
|$AM$|Data|Total number of adult males in removals|
|$AF$|Data|Total number of adult females in removals|
|$p(m|a)$|Parameter|Proportion of males in adults (or probability of being a male knowing it is an adult)|
|$CML_{i, t}$|Data|Cumulated number of male-like individuals counted in population $i$ in year $t$, which are assumed to all be adult males|
|$C_{i, t}$|Data|Cumulated number of individuals counted in population $i$ in year $t$|
|$p(a \cap m)_{i, t}$|Parameter|Proportion of adult males in population $i$ in year $t$|
|$p(a)_{i, t}$|Parameter|Proportion of adults in population $i$ in year $t$|
|$p(i)_{i, t}$|Parameter|Proportion of immatures in population $i$ in year $t$|
|$N_{i, t}$|Data|Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population)|
|$s_{i, t}$|Parameter|Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$|
|$r_{i, t}$|Parameter|Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per individual in year $t - 1$|
|$S_{i, t}$|Parameter|Number of adults in population $i$ in year $t$|
|$R_{i, t}$|Parameter|Number of recruits in population $i$ in year $t$|

\begin{equation}
p(m|a) \sim {\sf Beta}(AM, AF)
(\#eq:pm)
\end{equation}

\begin{align}
CML_{i, t} & \sim {\sf Binom}(p(a \cap m)_{i, t}, C_{i, t}) \notag \\
& \sim {\sf Binom}(p(m|a).p(a)_{i, t}, C_{i, t}) \notag \\
& \sim {\sf Binom}(p(m|a).(1 - p(i)_{i, t}), C_{i, t})
(\#eq:pic)
\end{align}

\begin{equation}
\begin{aligned}
s_{i, t} & = \frac{N_{i, t}}{N_{i, t - 1}}.(1 - p(i)_{i, t}) \\
r_{i, t} & = \frac{N_{i, t}}{N_{i, t - 1}}.p(i)_{i, t}
\end{aligned}
(\#eq:vr)
\end{equation}

\begin{equation}
\begin{aligned}
S_{i, t} & = N_{i, t}.(1 - p(i)_{i, t}) \\
R_{i, t} & = N_{i, t}.p(i)_{i, t}
\end{aligned}
(\#eq:VR)
\end{equation}

## Validation of the count-based method {#VAL}

The relevance of the method was assessed by testing the likelihood of the estimates of both vital rates. For adult survival rate, we checked if the values belonged to the expected interval $[0;1]$, and we compared the values to literature data on ducks that are accessible because adult survival is commonly assessed by capture-mark-recapture [@Lebreton2001]. Maximum adult survival is defined as a uniform distribution on $[0.7;1]$. The lower limit corresponds to the upper range of survival rates in literature for waterfowl species of similar weight, the upper limit was set to 1 because survival rates of long-life waterfowl species are very high [@Nichols1997; @Krementz1997; @Buxton2004; @Nichols1992]. Recruitment rate is more tricky to validate because it is not upper bounded as it is defined on $[0;\infty[$ and is neither explicitly informed in literature. Only some components of the recruitment rate are generally described, such as nesting rate, nesting success, clutch size, hatching success, early stage survival, but rarely survival between the juvenile stage and the first breeding season, which is necessary to inform recruitment rate [e.g. @Baldassarre2014]. We then developed an approach to estimate the maximum expected recruitment rate without exploitation and without negative density-dependence processes, and we checked if all values were equal or below this maximum recruitment rate. If the recruitment rate outputs of the model are significantly higher than maximum recruitment rate, it means that the method overestimates such rate, which reflects that a part of adult males are not detected during count surveys. A null or negative recruitment rate would mean that immature males moulted before the survey or/and the assumption on the stability of the adult sex ratio is overrated.

Practically, maximum recruitment rate is deduced by using its relationship to maximum growth rate and maximum adult survival rate, see Equation \@ref(eq:GR). This relationship comes from a simple reasoning for a closed population: population size in year $t$ is equal to the number of adults that survived over year $t - 1$ plus the offspring produced in year $t - 1$ that survived until the reproduction period of year $t$, i.e. the recruitment in year $t$. The growth rate of a population is thus the sum of adult survival rate plus recruitment rate [@Flint2015]. For an open population, adult survival and recruitment rates are confounded to adult and recruit migrations, respectively. This relationship becomes more complex if a species with delayed maturity is considered, see @Robertson2008.

|Name|Description|
|:--|:----------------------------|
|$N_{t}$|Number of adults in year $t$|
|$D_{t}$|Number of adults dead during year $t$|
|$R_{t}$|Number of recruits in year $t$|
|$s_{t}$|Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$|
|$r_{t}$|Recruitment rate, i.e. number of recruits in year $t$ produced per individual in year $t-1$|
|$\lambda_{t}$|Growth rate of the population between year $t-1$ and year $t$|

\begin{align} 
N_{t} & = N_{t-1} - D_{t-1} + R_{t} \notag \\
N_{t} & = N_{t-1} - (1 - s_{t}).N_{t-1} + r_{t}.N_{t-1} \notag \\
N_{t} & = s_{t}.N_{t-1} + r_{t}.N_{t-1} \notag \\
\frac{N_{t}}{N_{t-1}} & = s_{t} + r_{t} \notag \\
\lambda_{t} & = s_{t} + r_{t} \notag \\
r_{t} & = \lambda_{t} - s_{t}  
(\#eq:GR)
\end{align}

The maximum growth rate occurred for both populations during their expansion phase, when change in population size was the steepest. To estimate robust maximum population growth rates for both populations, we smoothed annual population growth rate over a consistent time period by using a linear regression on the logarithm scale, see Equation \@ref(eq:L). To do so, we discarded Great Britain data before 1972 since the size estimate of this population was noisy below 50 individuals (Figure \@ref(fig:count)). After reaching the threshold of 1,000 individuals, the Great Britain (GB) population growth showed a strong inflexion whereas no culling pressure was applied. This observation suggests that a negative density-dependence process might occur beyond 1,000 individuals and led to consider only the first sequence of the time series to infer on the maximum growth rate in Great Britain, i.e. 1972-1981. For the French (FR) population, the sequence without culling effort spread over the period 1994-1999.

>>>>

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Delayed dichromatism as a convenient tool to disentangle the effects of survival and productivity on the population dynamics in waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]
- Jean-François Maillard^[Office Français de la Biodiversité, France - jean-francois.maillard@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]
date: "`r Sys.setlocale('LC_TIME', 'English'); format(Sys.time(), '%d %B, %Y')`"
abstract: |
  Monitoring the number of individuals is by far the most popular strategy for investigating the environmental factors ruling population dynamics and for measuring the efficacy of management actions targeting population recovery, control or eradication. Monitoring numbers is unfortunately inefficient for assessing the mechanisms underlying demographic processes and more specifically to assess the extent to which population growth rate is affected by changes in adult survival rather than fluctuations in reproductive parameters. Usual methods to inform adult survival and productivity, such as capture-mark-recapture and analysis of game-hunting bag, rely on catching animals which suffers from three main drawbacks. Firstly, the selectivity of the catching methods can lead to a biased representation of the underlying population structure. Secondly, catching and releasing animals is quite invasive, which generates stress, and is thus an additional threat to often already endangered species. Thirdly, releasing may simply be impossible especially for invasive alien species that represent a threat to native taxa. In many waterfowl species, a sexual dichromatism is observed in adults whereas immatures of both sexes display a cover-up plumage similar to adult females. Using the case of alien Ruddy duck populations of Western Europe (a species displaying this pattern), we introduce a non-invasive method based on count data distinguishing plumage types that allows estimating the respective contribution of adult survival and productivity (also called recruitment rate) to population growth rate. We exemplify how, based on count surveys, changes in apparent male proportion (due to female-like immature males progressively acquiring adult male plumage) can be exploited to decompose population growth rate into adult survival and recruitment rates, greatly helping to understand the demographic impact of two different eradication strategies.

  **Keywords**: intrinsic growth rate - vital rate - reproductive success - fecundity - sex ratio - age ratio - Oxyura jamaicensis
output: 
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
    bookdown::word_document2:
      toc: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

Assessing population growth rate is a first step towards a better understanding of factors underlying the dynamics of natural populations [e.g. @Niel2005]. It is also crucial for measuring the efficacy of management actions potentially undertaken to help population recovery, control, or eradication [@Shea1998]. Among the approaches available to managers for reaching these goals, those relying on the monitoring of the number of individuals (counts) are by far the most popular [@Rintala2022]. In many instances, these methods enable investigating environmental factors underlying changes in population size, and hence help implementing mitigating actions [@Faillettaz2019]. Unfortunately, relying on the monitoring of population abundance alone generally hinders basic understanding of the demographic mechanisms underlying changes in population growth rates. More specifically, based on counts only, it is nearly impossible to assess the extent to which population growth rate is affected by changes in adult survival rather than to variations in productivity [@Austin2000], which is defined as the product of all reproductive parameters. To assess the relative sensitivity of population growth rate to factors affecting adult survival or reproductive parameters, demographers most often rely on the monitoring of individuals by capture-mark-recapture [CMR, @Lebreton1992]. In practice, however, one cannot always rely on the capture and release of individuals owing to the legal status of the species that precludes any additional disturbance (e.g. critically endangered species) or forbid the release of alive individuals (e.g. invasive species or pests). Furthermore, although highly efficient for assessing demographic parameters, capture-mark-recapture methods suffer from potential drawbacks such as being invasive (and may affect behaviour and hence survival or breeding success) and hardly affordable when time and money are limiting. Genetic monitoring is a non-invasive alternative to capture-mark-recapture, but it suffers from being costly and requires large samples [@Caniglia2011].

Most of the time therefore, managers do the best of a bad job using counts as a viable option for tracking population size changes and then population growth rate, in order to assess the relevance of management actions. Distinguishing immatures from adults in counts enables estimating adult survival rate, i.e. the proportion of breeders that survived over one year, and productivity/recruitment rate, i.e. the number of immatures produced per breeder that reach sexual maturity, which are, by definition, recruited in the breeding population. This additional effort makes it possible to assess the relative influence of each of these components on population growth rate. Alternatively, assessing the age structure in hunting bags has been used to infer the role of decreasing reproductive success in population declines in a number of games species including ducks and geese but suffers from intractable biases [@Fox2018]. Of course, the latter approach is not suited for protected/endangered species.

Here we exemplify how, in dimorphic species, delayed maturity of males can be used for estimating adult survival and recruitment rate by differentiating male-like from female-like individuals in repeated winter counts. We used two non-native Ruddy duck (*Oxyura jamaicensis*) populations introduced in Europe as a study model. As numerous duck species, Ruddy duck is dimorphic, with newly born males looking like females until the prenuptial moult, which occurs in late winter. As a result of this late prenuptial moult of immatures, the apparent proportion of males increases over the course of the wintering season. These changes in apparent proportions of males during this period are therefore directly related to the proportion of immatures into the populations and thus to the reproductive success of the previous breeding season. We estimated adult survival and recruitment rates and assessed the impact of two different eradication strategies deployed in Great Britain and in France, respectively.

@@

The Ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in the 1940's [@GutierrezExposito2020], a feral population began to establish in Great Britain with the first observed reproductive attempts in the wild in the 1960's (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5,000 individuals in the early 2000s. By the end of the 1980's, a feral population began to establish also on the continent, especially in France, supposedly owing to the arrival of individuals born in Great Britain. But contrary to what was observed in Great Britain, the Ruddy ducks did not spread much in France, with the vast majority of observations and breeding attempts clustering in the North-West of the territory. In France, during winter, almost no Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which greatly facilitates the monitoring of this population. Ruddy ducks are considered as a major threat to the endangered native White-headed duck (*Oxyura leucocephala*) of the South-Western Mediterranean, because of hybridization and thus an elevated risk for genetic pollution and genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, eradication measures were implemented in the late 1990's both in Great Britain and in France [@GutierrezExposito2020], and were followed by a European Ruddy duck eradication plan in 1999 [@Hughes1999] (Figure \@ref(fig:count)).

The trends and status of populations of both White-headed duck and Ruddy duck are being assessed through population censuses only. It is thus impossible to assess the factors that impair recovery of the former and the efficacy of the eradication program for the latter. In particular, based on censuses alone, one cannot assess the relative effects of changes in adult survival and recruitment rates on population growth rates which are prerequisites to identifying limiting factors.

```{r count, fig.cap = "(ref:count)"}
knitr::include_graphics("../Output/plot_paper_2_saved.png")
```

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism thereby young males acquire the typical colourful plumage of their kind, and hence can be distinguished from females, over the course of the interbreeding season, typically by mid-winter for the earliest individuals [@Baldassarre2014]. Delayed dichromatism generally explains the discrepancies in the estimates of apparent proportions of males counted in winter vs counted in spring, or vs males determined from individuals culled in winter (Figures \@ref(fig:flock) & \@ref(fig:pmal)).

Interestingly, delayed dichromatism enables estimating adult survival and recruitment/productivity by monitoring the seasonal evolution of apparent sex ratio and by making few and reasonable assumptions (see next section).

<!-- Interestingly, delayed dichromatism allows estimating the proportion of immature males, which is the difference between the proportion of female-like individuals between winter and spring. The proportion of all immature individuals is twice this value if we assume an even sex ratio at birth [@Bellrose1961; @Blums1996] and comparable mortality rates between immature females and immature males [@Ramula2018; @Wood2021]. The proportion of immatures can even be estimated only from winter counts by making the assumption of a known and unchanging adult sex ratio over the monitoring period [@Devineau2010], as developed inhere. By combining this proportion of immatures to reliable counts over two successive winters, both adult survival rate and recruitment rate can be estimated. -->

@@

```{r pmal, fig.cap = "(ref:pmal)"}
knitr::include_graphics("../Output/plot_paper_1_saved.png")
```

We monitored the demography of Ruddy duck populations in both Great Britain (GB) and France (FR) using exhaustive counts on the wintering grounds during the period between December 1 and January 31. In addition, apparent sex ratios in winter could be accessible from counts that distinguished female-like from male-like individuals in winter. Such counts were performed from 2006 to 2012 in Great Britain, which corresponded to a period of strong decline of the population (Figure \@ref(fig:count)). In France, a monitoring accounting for apparent sex was performed in 1999, 2001-2009, and 2012-2019. The French population observed a rapid growth during the first years of the monitoring and then stabilized from 2006 onwards as a result of a high culling pressure. In both countries, shooting was performed under the strict control of the authorities, meaning that the exact number of individuals culled was known. Therefore, despite the age ratio in birds shot was not ascertained before 2009 in France, it is possible to grossly qualify the culling pressure in three categories for both countries: "no culling" before 1999, "low culling" between 1999 and 2005, and "high culling" from 2006 (see Figure \@ref(fig:sup) in Supplementary Material).

## Inferring adult survival and recruitment rates from changes in population size and proportion of immatures

The proportion of immature males is the difference between the proportion of female-like individuals between winter and spring. The proportion of all immature individuals is twice this value if we assume an even sex ratio at birth [@Bellrose1961; @Blums1996] and comparable mortality rates between immature females and immature males [@Ramula2018; @Wood2021]. However, counts in spring suffer from inextricable issues, such as a differential detectability between females and males. Fortunately, the proportion of immatures can also be estimated only from winter counts by making the assumption of a known and unchanging adult sex ratio over the monitoring period [@Devineau2010].

The adult sex ratio is a prerequiste to estimate the proportion of immatures. No counts differentiating males from females were available during the breeding period to estimate adult sex ratios, but we determined the sex of each adult that was shot as part of the eradication programme both in Great Britain and in France. Nevertheless, the small number of culled individuals in France prevented from getting robust adult sex ratios. In Great Britain, the adult sex ratio estimated from the analysis of birds shot did not differ among months, so data from adults collected over the whole biological cycle were pooled to estimate adult sex ratios. The proportion of males did not statistically differ between the different years with more than 500 adults shot. This is consistent with previous findings showing that adult sex ratios in ducks are usually highly stable over the short-term [@Wood2021] but may vary over the long-term. Therefore, we pooled data from all adults shot as part as the control programme to estimate the adult sex ratio, see Equation \@ref(eq:pm).

Assuming the additive property of the binomial distribution, the proportion of immatures in a population can be deduced from the proportion of males in adults (alternatively the adult sex ratio) and the proportion of male-like individuals in the different count surveys in winter (assuming all these individuals are adult males), see Equation \@ref(eq:pic). By combining the population growth rate, which is derived from the year-to-year variation of an index of abundance of population size, to the proportion of immatures, one can derive adult survival and recruitment rates, see Equation \@ref(eq:vr) and Figure \@ref(fig:outline). A key concept is that if the absolute value of population size is known, numbers of adults and recruits can be estimated, see Equation \@ref(eq:VR).

| Variable name         | Description                                                                                                |
|:-|:-------|
| *Observation (input)* |                                                                                                            |
| $AM$                  | Total number of adult males shot as part of the eradication programme                                      |
| $AF$                  | Total number of adult females shot as part of the eradication programme                                    |
| $CML_{i, t}$          | Number of male-like individuals counted in population $i$ in year $t$                                      |
| $C_{i, t}$            | Number of individuals counted in population $i$ in year $t$                                                |
| $N_{i, t}$            | Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population) |
| *Parameter (output)*  |                                                                                                            |
| $p(m|a)$              | Proportion of males in adults (or probability of being a male knowing it is an adult)                      |
| $p(a \cap m)_{i, t}$  | Proportion of adult males in population $i$ in year $t$                                                    |
| $p(a)_{i, t}$         | Proportion of adults in population $i$ in year $t$                                                         |
| $p(re)_{i, t}$        | Proportion of immatures/recruits in population $i$ in year $t$                                             |
| $s_{i, t}$            | Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$                  |
| $r_{i, t}$            | Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per individual in year $t - 1$     |
| $S_{i, t}$            | Number of adults in population $i$ in year $t$                                                             |
| $R_{i, t}$            | Number of recruits in population $i$ in year $t$                                                           |

```{=tex}
\begin{equation}
p(m|a) \sim {\sf Beta}(AM, AF)
(\#eq:pm)
\end{equation}
```
```{=tex}
\begin{align}
CML_{i, t} & \sim {\sf Binom}(p(a \cap m)_{i, t}, C_{i, t}) \notag \\
& \sim {\sf Binom}(p(m|a).p(a)_{i, t}, C_{i, t}) \notag \\
& \sim {\sf Binom}(p(m|a).(1 - p(re)_{i, t}), C_{i, t})
(\#eq:pic)
\end{align}
```
```{=tex}
\begin{equation}
\begin{aligned}
s_{i, t} & = \frac{N_{i, t}}{N_{i, t - 1}}.(1 - p(re)_{i, t}) \\
r_{i, t} & = \frac{N_{i, t}}{N_{i, t - 1}}.p(re)_{i, t}
\end{aligned}
(\#eq:vr)
\end{equation}
```
```{=tex}
\begin{equation}
\begin{aligned}
S_{i, t} & = N_{i, t}.(1 - p(re)_{i, t}) \\
R_{i, t} & = N_{i, t}.p(re)_{i, t}
\end{aligned}
(\#eq:VR)
\end{equation}
```
```{r outline, fig.cap = "(ref:outline)"}
knitr::include_graphics("../Output/outline.png")
```

## Validating the method {#VAL}

Nevertheless, we developed an indirect approach to estimate the maximum expected recruitment rate without exploitation and assuming that the observed density levels did not affect components of the recruitment process. Estimated values above that maximum recruitment rate would reflect failure in detection of adult males. On the other hand, null or negative recruitment rates would reflect either that immature males had acquired the male plumage before the survey, or violation of the assumption about the stability of the adult sex ratio.

Practically, maximum recruitment rate is estimated from the difference between maximum growth rate and maximum adult survival rate, see Equation \@ref(eq:GR). This relationship comes from a simple reasoning for a closed population: population size in year $t$ is equal to the number of adults that survived over year $t - 1$ plus the offspring produced in year $t - 1$ that survived until the reproduction period of year $t$, i.e. the recruitment in year $t$ [@Flint2015]. For an open population, adult survival and recruitment rates are confounded to adult and recruit migrations, respectively, but this does not corrupt the equation. This relationship becomes more complex if a species with delayed maturity is considered, see @Robertson2008.

| Variable name | Description                                                                                 |
|:-|:---------|
| $N_{t}$       | Number of adults in year $t$                                                                |
| $D_{t}$       | Number of adults dead during year $t$                                                       |
| $R_{t}$       | Number of recruits in year $t$                                                              |
| $s_{t}$       | Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$   |
| $r_{t}$       | Recruitment rate, i.e. number of recruits in year $t$ produced per individual in year $t-1$ |
| $\lambda_{t}$ | Growth rate of the population between year $t-1$ and year $t$                               |

```{=tex}
\begin{align} 
N_{t} & = N_{t-1} - D_{t-1} + R_{t} \notag \\
N_{t} & = N_{t-1} - (1 - s_{t}).N_{t-1} + r_{t}.N_{t-1} \notag \\
N_{t} & = s_{t}.N_{t-1} + r_{t}.N_{t-1} \notag \\
\frac{N_{t}}{N_{t-1}} & = s_{t} + r_{t} \notag \\
\lambda_{t} & = s_{t} + r_{t} \notag \\
r_{t} & = \lambda_{t} - s_{t}  
(\#eq:GR)
\end{align}
```
Maximum adult survival was defined as a uniform distribution on the $[0.7;1]$ interval. The lower limit corresponded to the upper range of survival rates observed in waterfowl species of similar weight, and the upper limit was set to 1 because survival rates of long-lived waterfowl species are very high [@Nichols1997; @Krementz1997; @Buxton2004; @Nichols1992; @Koons2014].
